---
title: "Creating the training data"
date: "2022-12-22"
geometry: margin=1.5cm
output:
  word_document: default
  html_document: default
---
**Description:** This script is used to create the training data, combining the X an Y data created in the earlier steps

<br>

**The datasets used in this script are the following:**

* ny_x - This file contains the X data processed for all the tracts of New York MSA
* la_x - This file contains the X data processed for all the tracts of Los Angeles MSA
* ny_x - This file contains the Y labels for all the tracts of New York MSA
* la_y - This file contains the Y labels for all the tracts of Los Angeles MSA

<br>

**Using this script, the following datasets are created:**

* la_training_data.csv - This file is the training data for the Los Angeles MSA
* ny_training_data.csv - This file is the training data for the New York MSA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# clearing the workspace
rm(list=ls(all=TRUE))

# loading all the required libraries
library(readr, quietly = T)
library(data.table, quietly = T)
library(ROSE, quietly = T)
```

```{r}
# reading all the required files
ny_x = read.csv("ny_x.csv")
ny_y = read.csv("ny_y.csv")

la_x = read.csv("la_x.csv")
la_y = read.csv("la_y.csv")
```

```{r}
# viewing the datasets for New York MSA (X and y) to understand the layout of the datasets before combining them (same for Los Angeles MSA)
head(ny_x)
head(ny_y)
```

```{r}
# removing the 'X' column from the 'ny_y' dataset since it is not required and renaming the columns for readability
ny_y <- subset(ny_y, select = -c(X))
colnames(ny_y) <- c('AREAKEY', 'Gentrified')

# removing the 'X' column from the 'la_y' dataset since it is not required and renaming the columns for readability
la_y <- subset(la_y, select = -c(X))
colnames(la_y) <- c('AREAKEY', 'Gentrified')
```

```{r}
# merging the two files, by 'AREAKEY' (the unique tract identifier) and removing the column 'AREAKEY' since it is not required for training
ny_training_data <- merge(ny_x, ny_y, by = 'AREAKEY', how = 'inner')
ny_training_data <- subset(ny_training_data, select = -c(AREAKEY))

# removing null values (these null values emerged due to the ratios computed while creating the Y labels)
# since these are very small in number, we remove it
ny_training_data$Gentrified[is.na(ny_training_data$Gentrified)] = 0
ny_training_data$Gentrified <- as.factor(ny_training_data$Gentrified)

# merging the two files, by 'AREAKEY' (the unique tract identifier) and removing the column 'AREAKEY' since it is not required for training
la_training_data <- merge(la_x, la_y, by = 'AREAKEY', how = 'inner')
la_training_data <- subset(la_training_data, select = -c(AREAKEY))

# removing null values (these null values emerged due to the ratios computed while creating the Y labels)
# since these are very small in number, we remove it
la_training_data$Gentrified[is.na(la_training_data$Gentrified)] = 0
la_training_data$Gentrified <- as.factor(la_training_data$Gentrified)
```

```{r}
# creating csv files 
write_csv(ny_training_data, "ny_training_data.csv")
write_csv(la_training_data, "la_training_data.csv")
```