---
title: "Modeling for New York dataset"
date: "2022-12-22"
geometry: margin=1.5cm
output:
  word_document: default
  html_document: default
---
**Description:** This script is used to train, test and interpret the machine learning models for the tracts of the New York MSA.

<br>

**The datasets used in this script are the following:**

* ny_training_data.csv - This file is the training data for the New York MSA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# clearing the workspace
rm(list=ls(all=TRUE))

# loading all the required libraries
library(readr, quietly = T)
library(data.table, quietly = T)
library(dplyr, quietly = T)
library(caret, quietly = T)
library(randomForest, quietly = T)
library(smotefamily, quietly = T)
```

```{r}
# reading in the training dataset
ny_data = read.csv("ny_training_data.csv")

# converting the output variable to a factor  
ny_data$Gentrified <- as.factor(ny_data$Gentrified)
```

```{r}
# viewing the training dataset
head(ny_data)
```

```{r}
# normalizing the variables which are not in proportions (since proportions are already between 0 - 1)
minMax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

ny_data[c('WORKER30', 'AVWELIN0', 'AVSSI0', 'MDFAMY0', 'AVSOCS0', 'AVRETR0')] <- 
                                 lapply(ny_data[c('WORKER30', 'AVWELIN0', 'AVSSI0', 'MDFAMY0', 'AVSOCS0', 'AVRETR0')], minMax)
head(ny_data)
```
```{r}
# count of 0s and 1s of the output variable in the training data
print(table(ny_data$Gentrified))
```
 
```{r}
# randomly splitting the data into training and testing sets 
set.seed(12345)
tr <- sample.int(n = nrow(ny_data), size = floor(.8*nrow(ny_data)), replace = FALSE)
ny_data_train <- ny_data[tr, ]
ny_data_test <- ny_data[-tr, ]
```

```{r}
# count of 0s and 1s of the output variable in the training and testing sets
print(table(ny_data_train$Gentrified))
print(table(ny_data_test$Gentrified))
```
**Logistic Regression**

```{r}
# fitting a logistic regression model and printing the summary of the model
logistic_model <- glm(Gentrified ~ ., data = ny_data_train, family = "binomial")
summary(logistic_model)
```
```{r}
# obtaining the predictions on the testing set using the above model and printing the confusion matrix 
predlg1 <- predict(logistic_model, ny_data_test, type = 'response')
predlg1 <- ifelse(predlg1 > 0.5, 1, 0)
predlg1 <- as.factor(predlg1)
confusionMatrix(predlg1, ny_data_test$Gentrified, mode = 'everything', positive = '1')
```

```{r}
# sub-setting the training and testing sets with only the variables which are inferred as statistically significant from the logistic model
ny_data_train <- subset(ny_data_train, select = c(OLD0, WRKSM0, MDFAMY0, SHRHIP0, NONFAM0P, COMMUT20NWKH, WFEMP0P, VACHU0P, RNTOCC0P, Gentrified))
ny_data_test <- subset(ny_data_test, select = c(OLD0, WRKSM0, MDFAMY0, SHRHIP0, NONFAM0P, COMMUT20NWKH, WFEMP0P, VACHU0P, RNTOCC0P, Gentrified))
```

**Random Forest Classifier**

```{r}
# fitting a random forest classifier and printing the confusion matrix
rf1 <- randomForest(Gentrified ~ ., data=ny_data_train)
predrf1 <- predict(rf1, ny_data_test)
confusionMatrix(predrf1, ny_data_test$Gentrified, mode = 'everything', positive = '1')
```

**Re-sampling data to balance the two classes in the output variable**

```{r}
# using ADASYN (Adaptive Synthetic Sampling) technique to generate samples of the minority class
set.seed(1234)
ny_data_train_balanced <- ADAS(ny_data_train[,-ncol(ny_data_train)], ny_data_train$Gentrified)
ny_data_train_balanced <- ny_data_train_balanced$data
names(ny_data_train_balanced)[names(ny_data_train_balanced) == 'class'] <- 'Gentrified'

# converting the output variable to a factor 
ny_data_train_balanced$Gentrified <- as.factor(ny_data_train_balanced$Gentrified)

# count of 0s and 1s of the output variable in the new training set
table(ny_data_train_balanced$Gentrified)
```

**Random Forest Classifier (using the new training dataset)**

```{r}
# fitting a random forest classifier and printing the confusion matrix
rf2 <- randomForest(Gentrified ~ ., data=ny_data_train_balanced)
predrf2 <- predict(rf2, ny_data_test)
confusionMatrix(predrf2, ny_data_test$Gentrified, mode = 'everything', positive = '1')
```