---
title: "Y Data Preprocessing"
date: "2022-12-22"
geometry: margin=1.5cm
output:
  word_document: default
  html_document: default
---
**Description:** This script is used to create the Y labels (output variable) for all the tracts of New York and Los Angeles MSAs

<br>

**The datasets used in this script are the following:**

* Los_Angeles_2000_Y.csv - This file contains the variables mentioned in Table 1, for the year 2000 and the Los Angeles MSA
* Los_Angeles_2010_Y.csv - This file contains the variables mentioned in Table 1, for the year 2010 and the Los Angeles MSA
* New_York_2000_Y.csv - This file contains the variables mentioned in Table 1, for the year 2000 and the New York MSA
* New_York_2010_Y.csv - This file contains the variables mentioned in Table 1, for the year 2010 and the New York MSA

<br>

**Using this script, the following datasets are created:**

* la_y.csv - This file contains the Y labels for all the tracts of Los Angeles MSA
* ny_y.csv - This file contains the Y labels for all the tracts of New York MSA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# clearing the workspace
rm(list=ls(all=TRUE))

# import the required libraries
library(dplyr)
```

```{r}
#read in datasets
LA_2000 <- read.csv("Los_Angeles_2000_Y.csv")
LA_2010 <- read.csv("Los_Angeles_2010_Y.csv")
NY_2000 <- read.csv("New_York_2000_Y.csv")
NY_2010 <- read.csv("New_York_2010_Y.csv")
```

```{r}
# summary statistics for New York MSA variables in the year 2000
write.csv(summary(NY_2000[, -which(names(NY_2000) == 'AREAKEY')], digits = 2),"summary1.csv")
```

```{r}
# summary statistics for New York MSA variables in the year 2010
write.csv(summary(NY_2010[, -which(names(NY_2010) == 'AREAKEY')], digits = 2),"summary2.csv")
```

```{r}
# summary statistics for Los Angeles MSA variables in the year 2000
write.csv(summary(LA_2000[, -which(names(LA_2000) == 'AREAKEY')], digits = 2),"summary3.csv")
```

```{r}
# summary statistics for Los Angeles MSA variables in the year 2010
write.csv(summary(LA_2010[, -which(names(LA_2010) == 'AREAKEY')], digits = 2),"summary4.csv")
```

```{r}
#update column names for readability
colnames(LA_2000) <- c('Area Key','Total Population','Educated','Household Income','Median Home Value','Median Gross Rent')
colnames(LA_2010) <- c('Area Key','Total Population','Educated','Household Income','Median Home Value','Median Gross Rent')
colnames(NY_2000) <- c('Area Key','Total Population','Educated','Household Income','Median Home Value','Median Gross Rent')
colnames(NY_2010) <- c('Area Key','Total Population','Educated','Household Income','Median Home Value','Median Gross Rent')
```

```{r}
#create column for % college
LA_2000$"% College" <- (LA_2000$Educated/LA_2000$`Total Population`)*100
LA_2010$"% College" <- (LA_2010$Educated/LA_2010$`Total Population`)*100
NY_2000$"% College" <- (NY_2000$Educated/NY_2000$`Total Population`)*100
NY_2010$"% College" <- (NY_2010$Educated/NY_2010$`Total Population`)*100
```

```{r}
#change in CSA median for household income (LA)
CSA_hh_income_LA <- median(LA_2010$`Household Income`)-median(LA_2000$`Household Income`)

#change in CSA percentage for college educated (LA)
LA_2000_PC_educated <- (sum(LA_2000$Educated)/sum(LA_2000$`Total Population`))*100
LA_2010_PC_educated <-(sum(LA_2010$Educated)/sum(LA_2010$`Total Population`))*100
CSA_educated_LA <-LA_2010_PC_educated-LA_2000_PC_educated

#change in CSA median for gross rent (LA)
CSA_grent_LA <- median(LA_2010$`Median Gross Rent`)-median(LA_2000$`Median Gross Rent`)

#percent increase in CSA median for home value (LA)
CSA_increase_HV_LA <- ((median(LA_2010$`Median Home Value`)-median(LA_2000$`Median Home Value`))/(median(LA_2000$`Median Home Value`)))*100
```

```{r}
#change in CSA median for household income (NY)
CSA_hh_income_NY <- median(NY_2010$`Household Income`)-median(NY_2000$`Household Income`)

#change in CSA percentage for college educated (NY)
NY_2000_PC_educated <- (sum(NY_2000$Educated)/sum(NY_2000$`Total Population`))*100
NY_2010_PC_educated <-(sum(NY_2010$Educated)/sum(NY_2010$`Total Population`))*100
CSA_educated_NY <-NY_2010_PC_educated-NY_2000_PC_educated

#change in CSA median for gross rent (NY)
CSA_grent_NY <- median(NY_2010$`Median Gross Rent`)-median(NY_2000$`Median Gross Rent`)

#percent increase in CSA median for home value (NY)
CSA_increase_HV_NY <- ((median(NY_2010$`Median Home Value`)-median(NY_2000$`Median Home Value`))/(median(NY_2000$`Median Home Value`)))*100
```

```{r}
#create lists for columns in new dataframe (LA)
area_key_list_LA = LA_2000$`Area Key`
LA_hh_income_list = if_else((LA_2010$`Household Income`-LA_2000$`Household Income`)>CSA_hh_income_LA,1,0)
LA_educated_list = if_else((LA_2010$`% College`-LA_2000$`% College`)>CSA_educated_LA,1,0)
LA_grent_list = if_else((LA_2010$`Median Gross Rent`-LA_2000$`Median Gross Rent`)>CSA_grent_LA,1,0)
LA_HV_list = if_else((((LA_2010$`Median Home Value`-LA_2000$`Median Home Value`)/LA_2000$`Median Home Value`)*100)>CSA_increase_HV_LA,1,0)
```

```{r}
#create dataframe (LA)
LA_results <- data.frame(area_key_list_LA,LA_hh_income_list,LA_educated_list,LA_grent_list,LA_HV_list)
```

```{r}
#create lists for columns in new dataframe (NY)
area_key_list_NY = NY_2000$`Area Key`
NY_hh_income_list = if_else((NY_2010$`Household Income`-NY_2000$`Household Income`)>CSA_hh_income_NY,1,0)
NY_educated_list = if_else((NY_2010$`% College`-NY_2000$`% College`)>CSA_educated_NY,1,0)
NY_grent_list = if_else((NY_2010$`Median Gross Rent`-NY_2000$`Median Gross Rent`)>CSA_grent_NY,1,0)
NY_HV_list = if_else((((NY_2010$`Median Home Value`-NY_2000$`Median Home Value`)/NY_2000$`Median Home Value`)*100)>CSA_increase_HV_NY,1,0)
```

```{r}
#create dataframe (NY)
NY_results <- data.frame(area_key_list_NY,NY_hh_income_list,NY_educated_list,NY_grent_list,NY_HV_list)
```

```{r}
#create lists for final gentrification score
final_scores_LA = if_else((LA_results$LA_hh_income_list==1 & LA_results$LA_educated_list==1 & (LA_results$LA_grent_list==1 | LA_results$LA_HV_list==1)),1,0)
final_scores_NY = if_else((NY_results$NY_hh_income_list==1 & NY_results$NY_educated_list==1 & (NY_results$NY_grent_list==1 | NY_results$NY_HV_list==1)),1,0)
```

```{r}
#create final results dataframe
final_results_LA = data.frame(area_key_list_LA,final_scores_LA)
final_results_NY = data.frame(area_key_list_NY,final_scores_NY)
```

```{r}
#write dataframes to .csv files
write.csv(final_results_LA, "la_y.csv")
write.csv(final_results_NY, "ny_y.csv")
```
