---
title: "X Data Preprocessing"
date: "2022-12-22"
geometry: margin=1.5cm
output:
  word_document: default
  html_document: default
---
**Description:** This script is used to pre-process the X data (input variables) for all the tracts of New York and Los Angeles MSAs

<br>

**The datasets used in this script are the following:**

* Los_Angeles_X.csv - This file contains the variables mentioned in Table 3, for the year 2000 and the Los Angeles MSA
* New_York_X.csv - This file contains the variables mentioned in Table 3, for the year 2000 and the New York MSA

<br>

**Using this script, the following datasets are created:**

* la_x.csv - This file contains the X data processed for all the tracts of Los Angeles MSA
* ny_x.csv - This file contains the X data processed for all the tracts of New York MSA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# clearing the workspace
rm(list=ls(all=TRUE))

# loading all the required libraries
library(readr, quietly = T)
library(data.table, quietly = T)
library(ggplot2, quietly = T)
library(vtable, quietly = T)
```

```{r}
# reading the datasets
ny_df = read.csv("New_York_X.csv")
la_df = read.csv("Los_Angeles_X.csv")
```

```{r}
# changing the 'AREAKEY' (tract identifier variable) to factor from int data type
ny_df$AREAKEY <- as.factor(ny_df$AREAKEY)
la_df$AREAKEY <- as.factor(la_df$AREAKEY)

# removing rows which have zeros in all the columns
ny_df <- ny_df[rowSums(ny_df[,-1]) > 0,]
la_df <- la_df[rowSums(la_df[,-1]) > 0,]
```

```{r}
# calculating ratios for the required columns
ny_df['SHRAMI0'] = ny_df['SHRAMI0N']/ny_df['SHR0D']
la_df['SHRAMI0'] = la_df['SHRAMI0N']/la_df['SHR0D']

ny_df['SHRASN0'] = ny_df['SHRASN0N']/ny_df['SHR0D']
la_df['SHRASN0'] = la_df['SHRASN0N']/la_df['SHR0D']

ny_df['SHRHIP0'] = ny_df['SHRHIP0N']/ny_df['SHR0D']
la_df['SHRHIP0'] = la_df['SHRHIP0N']/la_df['SHR0D']

ny_df['NONFAM0P'] = ny_df['NONFAM0']/ny_df['NUMHHS0']
la_df['NONFAM0P'] = la_df['NONFAM0']/la_df['NUMHHS0']

ny_df['WNONFAM0P'] = ny_df['WNONFAM0']/ny_df['NUMHHS0']
la_df['WNONFAM0P'] = la_df['WNONFAM0']/la_df['NUMHHS0']

ny_df['COMMUT20NWKH'] = (ny_df['COMMUT20'] - ny_df['WKHOME0'])/ny_df['WRCNTY0D']
la_df['COMMUT20NWKH'] = (la_df['COMMUT20'] - la_df['WKHOME0'])/la_df['WRCNTY0D']

ny_df['WMEMP0P'] = ny_df['WMEMP0']/ny_df['WM16P0']
la_df['WMEMP0P'] = la_df['WMEMP0']/la_df['WM16P0']

ny_df['WFEMP0P'] = ny_df['WFEMP0']/ny_df['WF16P0']
la_df['WFEMP0P'] = la_df['WFEMP0']/la_df['WF16P0']

ny_df['VACHU0P'] = ny_df['VACHU0']/ny_df['TOTHSUN0']
la_df['VACHU0P'] = la_df['VACHU0']/la_df['TOTHSUN0']

ny_df['RNTOCC0P'] = ny_df['RNTOCC0']/ny_df['TOTHSUN0']
la_df['RNTOCC0P'] = la_df['RNTOCC0']/la_df['TOTHSUN0']
```

```{r}
# removing the columns not required (since these columns were converted to ratios and new columns were created)
ny_df <- subset(ny_df, select = -c(SHR0D, SHRAMI0N, SHRASN0N, SHRHIP0N, NONFAM0, WNONFAM0, NUMHHS0, COMMUT20, WKHOME0, WRCNTY0D, WMEMP0, WM16P0, WFEMP0, WF16P0, VACHU0, TOTHSUN0, RNTOCC0))

la_df <- subset(la_df, select = -c(SHR0D, SHRAMI0N, SHRASN0N, SHRHIP0N, NONFAM0, WNONFAM0, NUMHHS0, COMMUT20, WKHOME0, WRCNTY0D, WMEMP0, WM16P0, WFEMP0, WF16P0, VACHU0, TOTHSUN0, RNTOCC0))
```

```{r}
# replacing NANs introduced by the calculation of ratios in the above step with zeros
ny_df[is.na(ny_df)] <- 0
la_df[is.na(la_df)] <- 0
```

```{r}
# summary statistics for New York MSA variables in the year 2000
write.csv(summary(ny_df[, -which(names(ny_df) == 'AREAKEY')], digits = 2),"summary5.csv")
```

```{r}
# summary statistics for Los Angeles MSA variables in the year 2000
write.csv(summary(ny_df[, -which(names(ny_df) == 'AREAKEY')], digits = 2),"summary6.csv")
```

```{r}
# creating csv files 
write_csv(ny_df, "ny_x.csv")
write_csv(la_df, "la_x.csv")
```
